---
title: "Dr Semmelweis Analysis"
author: "Heng KongkeaOudong"
date: "January 08 2024"
output: github_document
---
# Setup and Context
## Introduction

Dr Ignaz Semmelweis was a Hungarian physician born in 1818 who worked in the Vienna General Hospital. In the past people thought of illness as caused by "bad air" or evil spirits. But in the 1800s Doctors started looking more at anatomy, doing autopsies and started making arguments based on data. Dr Semmelweis suspected that something was going wrong with the procedures at Vienna General Hospital. Semmelweis wanted to figure out why so many women in maternity wards were dying from childbed fever (i.e., [puerperal fever](https://en.wikipedia.org/wiki/Postpartum_infections)).

Today we will become Dr Semmelweis. We will step into Dr Semmelweis' shoes and analyse the same data collected from 1841 to 1849.

## The Data Source

Dr Semmelweis published his research in 1861. [full text with the original tables in German](http://www.deutschestextarchiv.de/book/show/semmelweis_kindbettfieber_1861), [English translation can be found here](http://graphics8.nytimes.com/images/blogs/freakonomics/pdf/the%20etiology,%20concept%20and%20prophylaxis%20of%20childbed%20fever.pdf).

## Libraries 
```{r}
library(dplyr)
library(tidyverse)  
library(lubridate)  
library(ggplot2)
library(readr)
library(stringr)
library(viridis)
library(tibbletime)
```
## Read the data
```{r}
monthly_df <- read_csv('monthly_deaths.csv')
yearly_df <- read_csv('annual_deaths_by_clinic.csv')
```
# Preliminary Data Exploration

Check out these two DataFrames.

  * What is the shape of yearly_df and monthly_df? How many rows and columns?
  * What are the column names?
  * Which years are included in the dataset?
  * Are there any NaN values or duplicates?
  * What were the average number of births that took place per month?
  * What were the average number of deaths that took place per month?
  
## Yearly dataframe
```{r}
yearly_df
```
```{r}
is.null(yearly_df)
```
```{r}
sum(duplicated(yearly_df))
```
```{r}
summary(yearly_df)
```
**Insight**: The `yearly_df` is cleaned, with no missing values or duplicates. The average annual births and deaths are 3153 and 223, respectively.

## Monthly dataframe
```{r}
monthly_df
```
```{r}
is.null(monthly_df)
```
```{r}
sum(duplicated(monthly_df))
```
```{r}
summary(monthly_df)
```
**Insight**: The `monthly_df` is also cleaned. In the monthly time frame, the average births and deaths are 267 and about 22, respectively.

# Percentage of Women Dying in Childbirth
How dangerous was childbirth in the 1840s in Vienna?

In comparison, the United States recorded 18.5 maternal deaths per 100,000 or 0.018% in 2013 [(source).](
https://en.wikipedia.org/wiki/Maternal_death#:~:text=The%20US%20has%20the%20%22highest,17.8%20per%20100%2C000%20in%202009)

```{r}
total_death <- sum(yearly_df$deaths)
total_birth <- sum(yearly_df$births)
death_rate <- (total_death/total_birth)*100
sprintf("The death rate in the 1840s in Vienna is: %.3f percent.", death_rate)
```
**Insight:** The general mortality rate of women in childbirth during the 1840s was a staggering 7.08%, which is significantly higher than the 2013 United States maternal death rate of only 0.018%. This implies that for every 100 women giving birth, 7 faced mortality each year.

# Visualise the Total Number of Births 🤱 and Deaths 💀 over Time
```{r}
ggplot(monthly_df, aes(x=date)) +
  geom_line(aes(y=births), linewidth = 1, color = "Blue") + 
  geom_line(aes(y=deaths), linewidth = 1, color = "Red") + 
  labs(title = "Total number of births and deaths monthly")+ 
  scale_y_continuous(
    
    # Features of the first axis
    name = "Births",
    
    # Add a second axis and specify its features
    sec.axis = sec_axis(~., name="Deaths")) +  
  theme(
    axis.title.y = element_text(color = "Blue", size=13),
    axis.title.y.right = element_text(color = "Red", size=13)
  ) 
```
<br/>**Insight:** What stands out here is that from 1841 to 1847, the number of monthly deaths appears to be relatively high. After 1847, there was a sharp increase in deaths initially, but this was followed by a significant decrease, leading to a consistently low death rate ever since which indicating that something might have occurred to cause this decline. It's also important to noted that after the significant drop in death rate, we observed that the number of births on the other hand skyrocketed.

# The Yearly Data Split by Clinic

Now let's look at the annual data instead.

  * Which clinic is bigger or more busy judging by the number of births?
  * Has the hospital had more patients over time?
  * What was the highest number of deaths recorded in clinic 1 and clinic 2?
```{r}
yearly_df
```
## Yearly birth split by clinic
```{r}
ggplot(yearly_df, aes(x=year, y = births, color = clinic)) + 
  geom_line(linewidth=1) +
  labs(title = "Yearly birth by clinc")+
  scale_color_manual(values=c("orange", "black"))
```
<br/> **Insight**: Overall, it is evident that there is an increase number of women are giving birth to children overtime and additionally, we see that Clinic 1 appears to be the busier than Clinic 2.
## Yearly deaths by clinic
```{r}
ggplot(yearly_df, aes(x=year, y = deaths, color = clinic)) + 
  geom_line(linewidth=1) +
  labs(title = "Yearly deaths by clinc") +
  scale_color_manual(values=c("darkorchid3", "goldenrod1"))
```
<br/>**Insight**: Clinic 2 has fewer deaths compared to Clinic 1, which is likely because Clinic 1 handles more patients. To better understand this, we should analyze the ratio of deaths to births at both clinics.

## Calculate the Proportion of Deaths at Each Clinic
```{r}
yearly_df <- yearly_df %>% 
  mutate(death_rate = round(deaths/births, 2))
yearly_df
```
```{r}
yearly_df %>% 
  group_by(clinic) %>% 
  summarise(average_death = round(mean(death_rate),2))
```
**Insight**: Now we can confidently say that, clinic 2 are safer considering the low average death rate of only 4% compare to clinic 1 10%.
```{r}
ggplot(yearly_df, aes(x=year, y = death_rate, color = clinic)) + 
  geom_line(linewidth=1) +
  labs(title = "Yearly death_rate by clinc")
```
**Insight:** It is evident that in 1842, the death rate was at its peak, with Clinic 1 recording a total death rate of over 15% and Clinic 2 at 7.5%. Further research on the internet suggests that during the 1840s, doctors and medical students were not as hygienic during autopsies, allowing diseases from the remains and particles of the corpse to transfer to women giving birth. This breakthrough allow Semmelweis to realised that the majority of the deaths in both clinics were not only cause by women giving birth but also transmitted disease.

# Major Breakthrough
After the discovery, in June 1847, Dr. Semmelweis mandated that every member of his medical staff begin cleaning their hands and instruments not only with soap and water but also with a chlorine solution after an autopsy.
## The Effect of Handwashing
```{r}
handwashing_start <- as.Date("1847-06-01", format = "%Y-%m-%d")
```
Adding a column called "pct_deaths" to `monthly_df` that has the percentage of deaths per birth for each row.
```{r}
monthly_df <- monthly_df %>% 
  mutate(pct_deaths = round(deaths/births, 2))
monthly_df
```

Creating two subsets from the `monthly_df` data: before and after Dr Semmelweis ordered washing hand.
```{r}
before_handwashing <- monthly_df[monthly_df$date < handwashing_start,]
after_handwashing <- monthly_df[monthly_df$date >= handwashing_start,]
```

Calculating the average death rate prior to June 1947 and the average death rate after June 1947.
```{r}
before_hw_rate = (sum(before_handwashing$deaths)/sum(before_handwashing$births))*100
after_hw_rate = (sum(after_handwashing$deaths)/sum(after_handwashing$births))*100
sprintf("Before the handwashing madatory the death rate was %.2f percentage.", before_hw_rate)
sprintf("After the handwashing madatory the death rate was %.2f percentage.", after_hw_rate)
```
**Insight**: Wow, the hand washing mandatory worked. The mortality rate dropped by roughly a factor of 5 from 10.53% to 2.15%.

# Calculate a Rolling Average of the Death Rate
```{r}
rolling_mean <- rollify(mean, window = 6)
before_handwashing <- before_handwashing %>% 
  mutate(
    pct_ma = rolling_mean(pct_deaths)
  )
combined_data <- bind_rows(
  before_handwashing %>% mutate(period = "Before handwashing"),
  after_handwashing %>% mutate(period = "After handwashing")
)
```
## Ploting the differences
```{r}
ggplot(combined_data, aes(x= date, y = pct_deaths)) +
  geom_line(data = before_handwashing, aes(y=pct_ma, color = "6m moving average"), linetype = "dashed", linewidth=1,)+
  geom_line(aes(color = period), linewidth = 0.7) + 
  scale_color_manual(values = c(
    "Before handwashing" = "black",
    "After handwashing" = "skyblue",
    "6m moving average" = "red"
  )) + 
  labs(title = "Percentage of Monthly Deaths over Time",
       x = "Date",
       y = "Percentage of Deaths",
       color = "Legend")+
  theme(aspect.ratio = 1/2)
```

