---
title: "Dr Semmelweis Analysis"
author: "Heng KongkeaOudong"
date: "December 30 2023"
output: github_document
---
# Setup and Context
## Introduction

Dr Ignaz Semmelweis was a Hungarian physician born in 1818 who worked in the Vienna General Hospital. In the past people thought of illness as caused by "bad air" or evil spirits. But in the 1800s Doctors started looking more at anatomy, doing autopsies and started making arguments based on data. Dr Semmelweis suspected that something was going wrong with the procedures at Vienna General Hospital. Semmelweis wanted to figure out why so many women in maternity wards were dying from childbed fever (i.e., [puerperal fever](https://en.wikipedia.org/wiki/Postpartum_infections)).

Today we will become Dr Semmelweis. We will step into Dr Semmelweis' shoes and analyse the same data collected from 1841 to 1849.

## The Data Source

Dr Semmelweis published his research in 1861. [full text with the original tables in German](http://www.deutschestextarchiv.de/book/show/semmelweis_kindbettfieber_1861), [English translation can be found here](http://graphics8.nytimes.com/images/blogs/freakonomics/pdf/the%20etiology,%20concept%20and%20prophylaxis%20of%20childbed%20fever.pdf).

## Libraries 
```{r}
library(dplyr)
library(tidyverse)  
library(lubridate)  
library(ggplot2)
library(readr)
library(stringr)
library(viridis)
```
## Read the data
```{r}
monthly_df <- read_csv('monthly_deaths.csv')
yearly_df <- read_csv('annual_deaths_by_clinic.csv')
```
# Preliminary Data Exploration

Check out these two DataFrames.

  * What is the shape of yearly_df and monthly_df? How many rows and columns?
  * What are the column names?
  * Which years are included in the dataset?
  * Are there any NaN values or duplicates?
  * What were the average number of births that took place per month?
  * What were the average number of deaths that took place per month?
  
## Yearly dataframe
```{r}
yearly_df
```
```{r}
is.null(yearly_df)
```
```{r}
sum(duplicated(yearly_df))
```
```{r}
summary(yearly_df)
```
**Insight**: The `yearly_df` is cleaned, with no missing values or duplicates. The average annual births and deaths are 3153 and 223, respectively.

## Monthly dataframe
```{r}
monthly_df
```
```{r}
is.null(monthly_df)
```
```{r}
sum(duplicated(monthly_df))
```
```{r}
summary(monthly_df)
```
**Insight**: The `monthly_df` is also cleaned. In the monthly time frame, the average births and deaths are 267 and about 22, respectively.

# Percentage of Women Dying in Childbirth
How dangerous was childbirth in the 1840s in Vienna?

  * Using the annual data, calculate the percentage of women giving birth who died throughout the 1840s at the hospital.

In comparison, the United States recorded 18.5 maternal deaths per 100,000 or 0.018% in 2013 [(source).](
https://en.wikipedia.org/wiki/Maternal_death#:~:text=The%20US%20has%20the%20%22highest,17.8%20per%20100%2C000%20in%202009)
